---
title: "Analysis of NOAA Storm Database"
author: "Nikhila Arkalgud"
date: "February 22, 2015"
output: html_document
---

Synopsis: 

We will use the NOAA Storm Database to exmaine the effects these disaster events have on lives. We will specifically try to analyze the effects on population health and economic consequences. From the results obtained we found that tornadoes had the highest effect on the population health, it had both the highest number of fatalities and injuries. The impact of tornadoes on population health were almost 2 to 3 times higher than the second highest event. Floods have resulted in the highest impact on the economy. Although economic damages due to floods are more than twice the second highest event(hurricane), it has not contributed to a high number of fatalities or injuries compared to tornadoes. The report below provides the steps taken to process this dataset and then provides the results.

Data Processing:

Download the Storm Data from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
Unzip the file and read the csv file into R.

```{r echo=TRUE,cache=TRUE}
stdata <- read.csv(bzfile("repdata-data-StormData.csv.bz2"))
```
We will check the dimentionality of this dataset
```{r echo=TRUE}
dim(stdata)
```
```{r echo=TRUE}
head(stdata)
```
We will find the total fatalities and injuries resulted from each disaster event type which is represented in this dataset under EVTYPE 

```{r echo=TRUE}
# aggregate the total fatalities for each evtype
ph_fatality <- aggregate(stdata$FATALITIES, by=list(stdata$EVTYPE), FUN=sum)
names(ph_fatality) <- make.names(c("evtype","fatality"))

# aggregate the total injuries for each evtype
ph_injury <- aggregate(stdata$INJURIES, by=list(stdata$EVTYPE), FUN=sum)
names(ph_injury) <- make.names(c("evtype","injury"))
```

We will now process these subsets to find the cummulative sum and cummulative fraction for both fatalities and injuries. This will help us isolate the events that cummulatively contribute to 95% of the total fatality and the events that cummulatively contribute to 95% of the total injuries.

```{r echo=TRUE}
# we will calculate the cummulative sum and find the cummulative fraction for each event
# this will help us identify the events that 
# cummulatively contribute to 95% of the total fatality

# sort the ph_fatality in desending order
phfo <- ph_fatality[order(-ph_fatality[,2]),]

#append a col cufrac to phfo
phfo$cufrac <- cumsum(phfo$fatality)/sum(phfo$fatality)
```

```{r echo=TRUE}
# we will calculate the cummulative sum and find the cummulative fraction for each event
# this will help us identify the events that 
# cummulatively contribute to 95% of the total injuries

# sort the ph_injury in desending order
phio <- ph_injury[order(-ph_injury[,2]),]

#append a col cufrac to phio
phio$cufrac <- cumsum(phio$injury)/sum(phio$injury)
```


We will also extract the total property damage and the crop damage resulted from each disaster event type.
Economic disaster for each event type can be measured using PROPDMG and PROPDMGEXP columns. The PROPDMGEXP column procides the magnitude, with 'K' representing thousand dollars, 'M' for millon dollars and 'B' for billion dollars. Similarly CROPDMG and CROPDMGEXP are used for estimating the crop damage.
We will add the property damages and crop damages to obtain the raw dollar amount of economic damage for each event type.

```{r echo=TRUE}
#append a column to replace 'K', 'M' and 'B' provided under CROPDMGEXP 
#with 1000,1000000,1000000000, empty rows are replaced by 1
#then multiply this value with the value provided under CROPDMG

# start by making all rows 1
stdata$cropDamage <- 1

#extract the K, M and B logical indices
kindices <- stdata$CROPDMGEXP == 'K'
mindices <- stdata$CROPDMGEXP == 'M'
bindices <- stdata$CROPDMGEXP == 'B'

# assign 1000, 1000000 and 1000000000 to replace the K, M and B
stdata$cropDamage[kindices] <- 1000
stdata$cropDamage[mindices] <- 1000000
stdata$crop[bindices] <- 1000000000

#multiply with CROPDMG column to get the actual economic damage for this event
stdata$cropDamage <- stdata$CROPDMG * stdata$cropDamage
```

```{r echo=TRUE}
#append a column to replace 'K', 'M' and 'B' provided under PROPDMGEXP 
#with 1000,1000000,1000000000, empty rows are replaced by 1
#then multiply this value with the value provided under PROPDMG

# start by making all rows 1
stdata$economicDamage <- 1

#extract the K, M and B logical indices
kindices <- stdata$PROPDMGEXP == 'K'
mindices <- stdata$PROPDMGEXP == 'M'
bindices <- stdata$PROPDMGEXP == 'B'

# assign 1000, 1000000 and 1000000000 to replace the K, M and B
stdata$economicDamage[kindices] <- 1000
stdata$economicDamage[mindices] <- 1000000
stdata$economicDamage[bindices] <- 1000000000

#multiply with PROPDMG column to get the actual economic damage for this event
#add the corresponding crop damage to this value
stdata$economicDamage <- (stdata$PROPDMG * stdata$economicDamage) + stdata$cropDamage

#aggregate the economic damage by each event type
ed <- aggregate(stdata$economicDamage, by=list(stdata$EVTYPE), FUN=sum)
names(ed) <- make.names(c("evtype","economicDamage"))

# sort the data by economicDamage in desending order
ed_ordered <- ed[order(-ed[,2]),]
```

```{r echo=TRUE}
# we will calculate the cummulative sum and find the cummulative fraction for each event
# this will help us identify the events that 
# cummulatively contribute to 95% of the total economic damage

#append a col cufrac to ed_ordered
ed_ordered$cufrac <- cumsum(ed_ordered$economicDamage)/sum(ed_ordered$economicDamage)
```

Results:

We will examine two keys effects that result due to the occurance of these events:

1. How does this affect the population health?

2. What are the economic consequences?

For each of these cases, examining the events that contribute to the 95% of the cummulative damage is a good measure to identify the most damaging events. We will display these events and their respective values. We will then plot the top 10 events for each of these categories.

1. How does this affect the population health?

We will analyze the number of fatalities and injuries that have occurred due to these events. This will help us analyze the effect these events have on the population health.

We will look at the events that contribute to 95% of the cummulative fatalites that results from these events.

```{r echo=TRUE}
# display the events that contribute to 90% of the total fatality
top95_f <- phfo[phfo$cufrac < 0.95,]

#total number of events
nrow(phfo)

#total number of events contributing to the 95% of total fatality
nrow(top95_f)

#list of these events contributing to the 95% of total fatality
top95_f
```

There are 32 events out of the total 985 events that contribute to 95% of the total number of fatalities.


The below bar chart shows top 10 events that contribute to the highest fatality

```{r echo=TRUE}
events <- top95_f$evtype[1:10]
labels <- paste(seq(1:10),events,sep=":")
ramp <- colorRamp(c("red","yellow","blue"))
barplot(top95_f$fatality[1:10], main = "Top 10 Events that have the highest number of fatalities", xlab = "Disaster Events", ylab = "Total number of fatalities", names.arg = seq(1:10), legend = labels, col = rgb(ramp(seq(0,1,length=10)),max=255))
```

From the barchart and also the table above we see that tornado has the highest number of fatalities.

```{r echo=TRUE}
top95_f[1,1:2]
```

Now we will examine the injuries that have resulted due to these disaster events.

We will look at the events that contribute to 95% of the cummulative injuries that results from these events.

```{r echo=TRUE}
# display the events that contribute to 95% of the total injuries
top95_i <- phio[phio$cufrac < 0.95,]

#total number of events
nrow(phio)

#total number of events contributing to the 95% of total injuries
nrow(top95_i)

#list of these events contributing to the 95% of total injuries
top95_i
```

There are 17 events out of the total 985 events that contribute to 95% of the total number of injuries.

The below bar chart shows top 10 events that contribute to the highest number of injuries.

```{r echo=TRUE}
events <- top95_i$evtype[1:10]
labels <- paste(seq(1:10),events,sep=":")
ramp <- colorRamp(c("red","yellow","blue"))
barplot(top95_i$injury[1:10], main = "Top 10 Events that have the highest number of injuries", xlab = "Disaster Events", ylab = "Total number of injuries", names.arg = seq(1:10), legend = labels, col = rgb(ramp(seq(0,1,length=10)),max=255))
```

From both the barchart and the table above we see that tornados also have also resulted in the highest number of injuries.

```{r echo=TRUE}
top95_i[1,1:2]
```

We can see that tornadoes have resulted in the highest number of fatalities and injuries, the numbers are almost 2-3 times higher than the second highest events leading it to have the highest impact on the population health.

2. What are the economic consequences? 

We will now examine the events that have the greatest economic consequences.
We will look at the property damage and the crop damage associated to each of these events. We will combine these two damages to obtain the economic damage for each event.

We will look at the events that contribute to 95% of the cummulative economic damage that results from these events.

```{r echo=TRUE}
# display the events that contribute to 95% of the total economic damage
top95_e <- ed_ordered[ed_ordered$cufrac < 0.95,]

#total number of events
nrow(ed_ordered)

#total number of events contributing to the 95% of total economic damage
nrow(top95_e)

#list of these events contributing to the 95% of total economic damage
top95_e
```

There are 18 events out of the total 985 events that contribute to 95% of the total economic damage.

The below bar chart shows top 10 events that contribute to the highest economic damage

```{r echo=TRUE}
events <- top95_e$evtype[1:10]

ed_in_billions <- top95_e$economicDamage[1:10]/1000000000

labels <- paste(seq(1:10),events,sep=":")

ramp <- colorRamp(c("red","yellow","blue"))

barplot(ed_in_billions, main = "Top 10 Events that have resulted in the highest economic damage in USD", xlab = "Disaster Events", ylab = "US Dollars in Billions", names.arg = seq(1:10), legend = labels, col = rgb(ramp(seq(0,1,length=10)),max=255))
```

From the barchart and the table above we see that Floods have resulted in the maximum economic damage.

```{r echo=TRUE}
top95_e[1,1:2]
```

Floods have resulted in the highest economic damage, which is a combined result of crop and property damages and resulted in high economic consequences, the US Dollars reported are more than twice the second highest event(hurricane). 

Interestingly, although floods have high economic consequences, they seem to have less effect on population health. Tornados in contrast are the highest in population health and have significant effects on economy as well, tornadoes are the third highest contributing event to economic damages.